---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: yes
---

##Introduction

Large amounts of data are now collected about our movements through out the day. There are a growing number of monitoring devices on the market. We are going to use one such set of data to demonstrate the principal of reproducible research. Below is a paper analyzing steps taken during a two month period, October and November 2012. The idea behind this paper is that everything needed to reproduce the results are all contained together. Writing papers in this way makes it much easier to reproduce and check results found by fellow data scientists and other researchers. With the added benefit that sharing data sets allows for new questions to be asked of existing data.

## Loading and preprocessing the data

As we have alredy discussed these data hold the number of steps taken over the two months. The data are held in a zip file in the same directory as our markdown file.

The activity monitoring data are made up of three columns: *steps*: containing the number of steps in a 5 minute interval; *date*: the date the steps were counted; and *interval*: the start time of the 5 minute interval in the format hhmm.

First load the R libraries we require for this analysis. Then we open the zip file and read the data into a variable called activity.

```{r, message=FALSE}
#Load the libraries we need
library(dplyr)
library(ggplot2)
#knitr::opts_chunk$set(fig.path = 'figure/')
options(scipen = 1, digits = 6)
```

```{r, results = "hide"}
#Make a connection to a zip file using unz
con <- unz(".\\activity.zip", filename = "activity.csv")

#Read the activity data from the zipped file into the dataframe activity
#Read the date column in the using the Date class
activity <- read.csv(con, 
                     colClasses = c("integer", "Date", "integer"))
```

Let's have a look at the data to get a feel for it.

```{r}
summary(activity)
head(activity)
str(activity)
```

## What is mean total number of steps taken per day?

The total number of steps a day can be found by adding up all the steps from a particular date. The following section of code uses the dplyr package in *r* to group all the data from each day together before adding up the steps.

```{r}

stepsperday <- activity %>%
    
    #Group together the data for each day
    group_by(date) %>%
    
    #Populate a new column StepsPerDay with the total steps
    #Ignore missing values
    summarize(StepsPerDay = sum(steps, na.rm = TRUE))
```

The summarised data looks like the following:

```{r}
head(stepsperday)
```

Now we find the average number of steps and the median number of steps across the two months.

```{r}
meanSteps <- mean(stepsperday$StepsPerDay)
medianSteps <- median(stepsperday$StepsPerDay)
```

Let's plot a histogram showing the frequency of particular numbers of steps per day. The plot also shows both the median and mean steps per day.

```{r}
#Save some labels to be used in the graph
meanLabel <- paste("Average steps per day: ",round(meanSteps,0))
medianLabel <- paste("Median steps per day: ", round(medianSteps))

g <- ggplot(stepsperday, aes(x = stepsperday$StepsPerDay))

#Add a histogram of the data in intervals of 500 steps
g <- g + geom_histogram(binwidth = 500)

#Add title along with x and y labels
g <- g + labs(title="Distribution of steps per day") +
    labs(x="Number of steps", y="Count")

#Show the mean and a label
g <- g + geom_vline(xintercept=meanSteps)
g <- g + geom_text(x = meanSteps + 500, y = 9,
                   hjust=0, vjust=0,
                   size = 4,
                   label = meanLabel)

#Show the median with a label
g<- g + geom_vline(xintercept=medianSteps)
g <- g + geom_text(x = medianSteps + 500, y = 8.25,
                   hjust=0, vjust=0,
                   size = 4,
                   label = medianLabel)

#plot the data
g
```
As you can see from the graph the average steps per day is `r round(meanSteps, 0)` and the median is `r round(medianSteps, 0)`.

## What is the average daily activity pattern?

The number of steps varies throughout the day. To get a feel for how many steps are taken at different times through the day we are going to take a look at the averge number of steps at each time period.

```{r}
dailyActivity <- activity %>%
    
    #Group together each time period through out the day
    group_by(interval) %>%
    
    #Average the number of steps taken at each time over the two months
    #Ignore missing values
    summarise(avgSteps = mean(steps, na.rm = TRUE))
```
The daily number of steps at different times of the day.

```{r}
head(dailyActivity)
summary(dailyActivity)
```

Here is a plot of the steps over the day.

```{r}
#Find the maximum number of steps and at what time is occurs
maxsteps <- max(dailyActivity$avgSteps)
maxstepstime <- filter(dailyActivity, avgSteps == maxsteps)
maxLabel <- paste("Maximum steps in 5 minutes is ", 
                  round(maxsteps, 1),
                  " at ", maxstepstime$interval)

g <- ggplot(dailyActivity, aes(x=interval, y=avgSteps))

#Plot the data as a time series
g <- g + geom_line()

#Add titles and labels
g <- g + labs(title = "Steps through out the day") +
    labs(x = "Time in of day in hours and minutes", 
         y = "Number of steps per 5 minutes")

#Show the average maximum steps and the time it occurs
g <- g + geom_vline(xintercept=maxstepstime$interval)
g <- g + geom_hline(yintercept=maxsteps)
g <- g + geom_text(x = maxstepstime$interval + 60,
                   y = maxsteps - 15,
                   hjust=0, vjust=0,
                   size = 4,
                   label = maxLabel)

#plot the graph
g
```

The plot shows that the maximum number of steps occurs at `r round(maxstepstime[1], 0)` and is on average `r round(maxsteps, 0)`.

## Imputing missing values

Ignoring missing values might not give us an accurate picture of what is going on. Ignoring missing values might introduce some bias into the results.

We impute the missing values using the average number of steps at each time period that we generated above.

```{r}
#Find the location of the missing values
missingSteps <- is.na(activity$steps)
#How many values are we missing?
missingValues <- sum(is.na(activity$steps))
```

There are a total of `r missingValues` missing values from our data set. Below we replace the missing values with the average steps for that time of day, calculated in the previous section.

```{r}
#Create a vector of replacement step values matching the 
#average step count for that time of day.
replaceSteps <- activity %>%
    
    #select only the missing entries
    filter(missingSteps) %>%
    
    #match the timestamp with the average for that time
    left_join(dailyActivity) %>%
    
    #keep only the vector of replacement averages
    select(avgSteps)

#Collect all the step values together with the missing values replaced
imputeSteps <- replace(activity$steps, missingSteps, replaceSteps$avgSteps)

#Create our new dataset then update the steps to the imputed values
activityImpute <- activity
activityImpute$steps <- imputeSteps
```

Now when we see the first few lines of our newly imputed dataset *activityImpute* the missing values are filled in.

```{r}
head(activityImpute)
```

Imputing the missing data changes the histogram of steps per day as we can see below.

```{r}
#Mean number of steps per day
stepsperday <- activityImpute %>%
    
    #Group the days together
    group_by(date) %>%
    
    #Add the steps for the day using the imputed data
    summarize(StepsPerDay = sum(steps, na.rm = TRUE))

#Find the average and median values
meanSteps <- mean(stepsperday$StepsPerDay)
medianSteps <- median(stepsperday$StepsPerDay)

#Start building the plot
meanLabel <- paste("Average steps per day: ",round(meanSteps,0))
medianLabel <- paste("Median steps per day: ", round(medianSteps))

g <- ggplot(stepsperday, aes(x = stepsperday$StepsPerDay))

#Generage a histogram of the frequency of different daily totals
g <- g + geom_histogram(binwidth = 500)

#Add some labels
g <- g + labs(title="Distribution of steps per day with missing values imputed") +
    labs(x="Number of steps", y="Count")

#Show the mean
g <- g + geom_vline(xintercept=meanSteps)
g <- g + geom_text(x = meanSteps + 500, y = 9,
                   hjust=0, vjust=0,
                   size = 4,
                   label = meanLabel)

#Show the median
g<- g + geom_vline(xintercept=medianSteps)
g <- g + geom_text(x = medianSteps + 500, y = 8.25,
                   hjust=0, vjust=0,
                   size = 4,
                   label = medianLabel)

#plot
g
```

The mean and median rounded to zero decimal places are the same at `r round(meanSteps, 0)`. We can see that the imputed data is less skewed toward the lower end of steps per day and the median and mean are close together.


## Are there differences in activity patterns between weekdays and weekends?
